---
title: "Problem Set 1"
author: "Jennifer Reeve"
date: "October 1, 2018"
output: pdf_document
header-includes:
  \usepackage{chemformula}
  \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Due Oct 15, show all work and state any assumptions.

```{r table inputs, echo=FALSE}
system<-data.frame(mineral=c("an","gr","ky","q"),
                   thermal_expanse=c(14.3,30.0,11.2,8.0),
                   compress=c(13.0,7.9,1.9,5.9),
                   a=c(0.3914,0.7286,.3039,.0979),
                   b=c(1.2556,-4.0986,-1.3390,-.3350)*10^-5,
                   c=c(-3036.2,-3128.0,-895.2,-636.2),
                   d=c(-2.5832,-6.0774,-2.9040,-.7740),
                   S=c(199.3,256.0,82.3,41.5)*10^-3,
                   V=c(10.079,12.535,4.414,2.269))
```

## Question 1: Thermodynamic Units
Convert thermodynamic units for volume (J/bar) into something more intuitive. Use the attached data table (Table 7 from Holland and Powell 1990) to determine the molar volume of *an* (\ch{CaAl2Si2O8}), $\bar{V_{an}}$ at 1 bar, 25C. Express the volume of one mole of *an* in terms of the volume of a common familiar object.

$\bar{V}=\frac{J}{bar}$

$J=\frac{kg*m^2}{s^2}$

$bar=100,000 Pa$

$Pa=\frac{kg}{m*s^2}$

$\bar{V}=\frac{\frac{kg*m^2}{s^2}}{100,000\frac{kg}{m*s^2}}=\frac{1}{100,000}m^3$

The molar volume of *an* at standard pressure and temperature (1 bar, 25C) is $10.079 \frac{J}{bar}$ or $1.0079*10^{-4} m^3$. To put this in perspective, a gallon is $3.79*10^{-3} m^3$ so 37.6 molar volumes of *an* would fit in one gallon.

```{r question 1, echo=FALSE,include=FALSE,message=FALSE}
gallon<-3.79e-3
an<-1.007e-4
aningallon<-gallon/an
```

## Question 2: Compressibility and thermal expansion
One expression of the molar volume ($\bar{V_j}$) of a pure substance ($j$) as a function of pressure and temperature is (Holland and Powell 1990): $\bar{V_j}(P,T)=\bar{V_j}^{P_0,T_0}+\alpha V * (T-T_0)-\beta V*(P-P_0)$

Use this expression to show that $(\frac{\partial \bar{V_j}}{\partial T})_P$ and $(\frac{\partial \bar{V_j}}{\partial P})_T$ are constant. [Note that $\alpha V$ and $\beta V$ are each single constant parameters in this expression.]

For this expression, all of the terms in blue below are constants under all circumstances, the terms in red are constants for $(\frac{\partial \bar{V_j}}{\partial T})_P$ and the terms in green are constants for $(\frac{\partial \bar{V_j}}{\partial P})_T$

$$\bar{V_j}(P,T)=\color{blue}{\bar{V_j}^{P_0,T_0}}+\color{blue}{\alpha V} * (\color{green}{T}-\color{blue}{T_0})-\color{blue}{\beta V}*(\color{red}{P}-\color{blue}{P_0})$$

So if we take the derivative of the molar volume with respect to temperature holding pressure constant then all terms in blue or red are constants and thus their derivatives are $0$, leaving us with only the derivative of the middle term:

$$(\frac{\partial \bar{V_j}}{\partial T})_P=0+\frac{d\alpha V T}{dT} -0-0=\alpha V$$

And for the derivative of the molar volume with respect to pressure holding temperature constant all of the terms in blue or green are constants so:

$$(\frac{\partial \bar{V_j}}{\partial P})_T=0+0-\frac{d\beta VP}{dP}+0=-\beta V$$

And as established above, both $\alpha V$ and $-\beta V$ are constants.

## Question 3: Equations of state
Use the equation of state for volumes of pure substances from Holland and Powell (1990) and the attached data table (Table 7 from Holland and Powell 1990) to calculate how (i) the molar volume, (ii) thermal expansion, and (iii) compressibility of *an* change along a typical geothermal gradient ($\approx \frac{25C}{km}$) between surface conditions (1 bar and 25C) and those at 35km depth. Plot your results.

```{r question 3}


```

## Question 4: Third condition of heterogeneous equilibrium for pure substances
Calculate and plot the pressure-temperature conditions of the *gr-ky-qz-an* equilibrium from the data in the attached table (Table 7 from Holland and Powell 1990). Only consider $T>500C$ and $P<25kbar$. [Note that Holland and Powell 1990 use the symbol *q* for \ch{SiO2}.]

```{r question 4}

```

## Question 5: Thermodynamic simplifications
Your plot in Question 4 should be very nearly a straight line in $P-T$ space. Discuss how this result might enable you to simplify the expression you used to make your calculations in Question 4. [Hint: are there any terms that you can ignore? Why?]
