---
title: "Problem Set 2"
author: "Jennifer Reeve"
date: "November 5, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
r_jmolk=8.31441
r_barcm3molk=83.1441
r_calmolk=1.987
```


## Question 1
Calculate $\bar{V}_{H_2O}$ at $T=800C$ and $P=1000 bars$ from figure of the compressibility factor as a function of $P$ and $T$. For reference $P_{crit}=220.4 bar$ and $T_{crit}=647.3 K$ for pure H2O.

Reduced pressure:
```{r}
p=1000
p_crit=220.4
press_reduced=p/p_crit
print(press_reduced)
```

Reduced temperature:
```{r}
t=800+273.15
t_crit=647.3
t_reduced=t/t_crit
print(t_reduced)
```

From the figure: $Z=0.88$

$Z\equiv\frac{P\bar{V}}{RT}$ so $\bar{V}=\frac{ZRT}{P}$

```{r}
z=0.88
molar_volume=z*t*r_barcm3molk/p
print(molar_volume)
```

We get $78.52 cm^3$ as the molar volume of pure water at 800C and 1000 bars.

## Question 2
Calculate $\bar{V}_{H_2O}$ and $f_{H_2O}^{P,T}$ at $T=800C$ and $P=1000 bars$ from the modified Redlich-Kwong equation of state. Use the following formulations for the attraction coefficient ($a$) and the excluded volume ($b$) from Bowers and Helgeson (1983) where T is in C.

The modified Redlich-Kwong equation of state is: $P=\frac{RT}{\bar{V}-b}-\frac{a}{\sqrt{T}\bar{V}(\bar{V}+b)}$

We are given $P$ and $T$ and thus can rearrange this to be: $0=P-\frac{RT}{\bar{V}-b}+\frac{a}{\sqrt{T}\bar{V}(\bar{V}+b)}$ and now we just need to find a $\bar{V}$ that satisfies this.

```{r}
t_c=800
a=10^6*(111.3+50.7*exp(-0.983*10^(-2)*t_c))
b=14.6
v_poss<-c(1,10,20,30,40,50,60,70,80,90,100)
outcomes<-p-(r_barcm3molk*t/(v_poss-b))+a/(sqrt(t)*v_poss*(v_poss+b))
print(outcomes) # know we know that the actual molar volume is somewhere in the range of 70 to 80 cm3 (the apparent minimum between in 10 and 20 is asymptotic in nature and should be ignored)
v_poss<-c(70,71,72,73,74,75,76,77,78,79,80)
outcomes<-p-(r_barcm3molk*t/(v_poss-b))+a/(sqrt(t)*v_poss*(v_poss+b))
print(outcomes) # getting more specific, it is somewhere between 72 and 73 cm3
v_poss<-c(72,72.1,72.2,72.3,72.4,72.5,72.6,72.7,72.8,72.9,73)
outcomes<-p-(r_barcm3molk*t/(v_poss-b))+a/(sqrt(t)*v_poss*(v_poss+b))
print(outcomes) # let's do one last round between 72.7 and 72.8
v_poss<-c(72.7,72.71,72.72,72.73,72.74,72.75,72.76,72.77,72.78,72.79,72.8)
outcomes<-p-(r_barcm3molk*t/(v_poss-b))+a/(sqrt(t)*v_poss*(v_poss+b))
print(outcomes)
```

The modified Redlich-Kwong equation of state predicts a molar volume of water at $T=800$ and $P=1000$ of approximately $72.725 cm^3$, slightly lower than the volume prediced by the figure estimate.


## Question 3
We are looking for equilibrium conditions in P,T space, so we are going to try to find the conditions where $\Delta\bar{G}=0$

$$0=\Delta\bar{G}^\circ(P,T)=\Delta(\bar{H}^\circ)^{1,298}-T_{eq}\Delta(\bar{S}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\Delta\bar{C_P}^\circ dT-T_{eq}\displaystyle\int^{T_{eq}}_{298}\frac{\Delta\bar{C_P}^\circ}{T}dT+\displaystyle\int^{P_{eq}}_1\Delta\bar{V}^\circ dP$$

But unlike before, we need to break the volume term down into the solid and fluid components as these behave differently:$0=\Delta\bar{G}^\circ(P,T)=\Delta(\bar{H}^\circ)^{1,298}-T_{eq}\Delta(\bar{S}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\Delta\bar{C_P}^\circ dT-T_{eq}\displaystyle\int^{T_{eq}}_{298}\frac{\Delta\bar{C_P}^\circ}{T}dT+\displaystyle\int^{P_{eq}}_1\Delta\bar{V_s}^\circ dP+\displaystyle\int^{P_{eq}}_1\nu_f\bar{V_f}dP$

From the last problem set we know that the slope doesn't change with $P$ or $T$ so:

$\Delta(\bar{H}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\Delta\bar{C_P}^\circ dT\approx\Delta(\bar{H}^\circ)^{1,T}$

$\Delta(\bar{S}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\frac{\Delta\bar{C_P}^\circ}{T}dT\approx \Delta(\bar{S}^\circ)^{1,T}$

$\frac{dP}{dT}=\frac{\Delta\bar{S}}{\Delta\bar{V}}$

Once we substitute these in we have:

$\Delta(\bar{G}^\circ)^{P,T}=\Delta(\bar{H}^\circ)^{1,T}-T\Delta(\bar{S}^\circ)^{1,T}+\Delta\bar{V_s}^\circ(P-P_r)+\displaystyle\int^{P_{eq}}_1\nu_f\bar{V_f}^\circ dP$

Now we need to deal with the volume term for our fluids. So our only complication is fugacity. For an ideal gas: $(f_i^\circ)^{P,T}=P$ so $\displaystyle\int^P_1\bar{V_f}^\circ dP=\nu_fRTln(f_f^\circ)^{P,T}$

Using a table of fugacities how can we do this?

$$\Delta(\bar{H}^\circ)^{1,T}-T\Delta(\bar{S}^\circ)^{1,T}+\Delta\bar{V_s}^\circ(P-P_r)+\nu_fRTln(f_f^\circ)^{P,T}=0$$

$\nu_f=1$ so we can ignore that.

Now the plan of attack (shown explicitly once in the code):

1. pick a $T$ to start
2. pick a $P$ which sets $(f_i^\circ)^{P,T}$ from the table
3. evaluate $\Delta(\bar{G}^\circ)^{P,T}$
4. Is $\Delta(\bar{G}^\circ)^{P,T}>0 \rightarrow$ choose a lower pressure, if $\Delta(\bar{G}^\circ)^{P,T}<0\rightarrow$ choose a higher pressure
5. iterate

```{r}
h=24184 #cal
s=40.817 #cal K-1
delta_barv_solid=-0.113 #cal bar-1
p_ref<-1
```

First we have to figure out what temperature to start at. We need to find the lowest temperature where the lowest pressure and fugacity give a negative $(\Delta\bar{G}^\circ)^{P,T}$

```{r}
t<-200+273.15
p<-100
f<-15
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this one is really high, but it's already at the lowest pressure we have data for so we are going to skip this temperature
t<-300+273.15
f<-68
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #still not an option
t<-400+273.15
f<-87
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #still not an option
t<-500+273.15
f<-92
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # woo we have a negative! Let's check the temperature above this to see if it was negative too:
t<-480+273.15
f<-92
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # nope, so we'll start at T=500
```

Alright so the lowest temperature we care about is 500C, so let's start figuring out the pressure conditions that provide us with equilibrium:

```{r}
t<-500+273.15
p<-100
f<-92
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #we checked this before, now let's check the next pressure
p<-200
f<-170
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 100 and 200 bars
#onto the next temperature
t<-520+273.15
p<-200
f<-173
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-300
f<-240
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 200 and 300 bars
t<-540+273.15
p<-300
f<-245
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-400
f<-305
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 300 and 400 bars
t<-560+273.15
p<-500
f<-369
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-600
f<-417
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 500 and 600 bars
t<-580+273.15
p<-900
f<-562
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-1000
f<-601
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 900 and 1000 bars
t<-600+273.15
p<-1400
f<-794
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-1500
f<-834
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 1400 and 1500 bars
t<-620+273.15
p<-2000
f<-1109
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-2100
f<-1152
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 2000 and 2100 bars
t<-640+273.15
p<-2800
f<-1603
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-2900
f<-1663
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 2800 and 2900 bars
t<-660+273.15
p<-3600
f<-2236
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-3700
f<-2312
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 3600 and 3700 bars
t<-680+273.15
p<-4400
f<-3035
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-4500
f<-3130
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 4400 and 4500 bars
t<-700+273.15
p<-5300
f<-4162
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-5400
f<-4285
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 5300 and 5400 bars
t<-720+273.15
p<-6200
f<-5583
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-6300
f<-5739
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 6200 and 6300 bars
t<-740+273.15
p<-7200
f<-7560
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-7300
f<-7760
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 7200 and 7300 bars
t<-760+273.15
p<-8100
f<-9834
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-8200
f<-10082
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 8100 and 8200 bars
t<-780+273.15
p<-9100
f<-12822
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-9200
f<-13130
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 9100 and 9200 bars
t<-800+273.15
p<-10000
f<-16214
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # even the highest pressure is negative, so this one is a no-go
t<-c(500,520,540,560,580,600,620,640,660,680,700,720,740,760,780)
t<-t+273.15
lower_p<-c(100,200,300,500,900,1400,2000,2800,3600,4400,5300,6200,7200,8100,9100)
upper_p<-c(200,300,400,600,1000,1500,2100,2900,3700,4500,5400,6300,7300,8200,9200)

df<-data.frame(t,lower_p,upper_p)

```

Now that we have our temperatures and pressure ranges, let's plot it:
```{r}
ggplot(df,aes(x=t))+geom_point(mapping=aes(y=lower_p),color='blue')+geom_point(mapping=aes(y=upper_p),color='red')
```
## Question 4
Following the same approach as in Question 3, but with different constants and fugacity table.

```{r}
h=23019 #cal
s=38.334 #cal k-1
delta_barv_solid=-0.471 #cal bar-1

t<-350+273.15
p<-1000
f<-1096
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # this is positive at the lowest pressure, so let's start at a higher temperature

t<-700+273.15 # starting here b/c that is the first temperature at which the lowest pressure is negative
p<-1000
f<-1304
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-2000
f<-3495
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-750+273.15
p<-2000
f<-3483
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-3000
f<-7713
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-800+273.15
p<-3000
f<-7555
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-4000
f<-14509
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-850+273.15
p<-4000
f<-13999
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-5000
f<-24268
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-900+273.15
p<-6000
f<-37399
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-7000
f<-58312
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-950+273.15
p<-8000
f<-83221
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-9000
f<-125468
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-1000+273.15
p<-9000
f<-117699
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-10000
f<-177096
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))

t<-1050+273.15
p<-9000
f<-111095
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f))
p<-10000
f<-168130
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # the highest pressure is still negative here, so we can't find the lower and upper bounds and won't use this temperature or anything higher

t<-c(700,750,800,850,900,950,1000)
lower_p<-c(1000,2000,3000,4000,6000,8000,9000)
upper_p<-c(2000,3000,4000,5000,7000,9000,10000)
df<-data.frame(t,lower_p,upper_p)
```

Now let's plot our data:
```{r}
ggplot(df,aes(x=t))+geom_point(mapping=aes(y=lower_p),color='blue')+geom_point(mapping=aes(y=upper_p),color='red')
```
## Question 5
Higher pressures than equilibrium pressures yield positive Gibbs free energy while lower pressures than the equilibrium pressures yield negative Gibbs free energy. This means that at higher pressures the reaction will not proceed in the direction written, but rather the reverse reaction is the reaction that will occur spontaneously. At lower pressures the reaction as written is the reaction that will proceed spontaneously.