---
title: "Problem Set 2"
author: "Jennifer Reeve"
date: "November 5, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
r_jmolk=8.31441
r_barcm3molk=83.1441
r_calmolk=1.987
```


## Question 1
Calculate $\bar{V}_{H_2O}$ at $T=800C$ and $P=1000 bars$ from figure of the compressibility factor as a function of $P$ and $T$. For reference $P_{crit}=220.4 bar$ and $T_{crit}=647.3 K$ for pure H2O.

Reduced pressure:
```{r}
p=1000
p_crit=220.4
press_reduced=p/p_crit
print(press_reduced)
```

Reduced temperature:
```{r}
t=800+273.15
t_crit=647.3
t_reduced=t/t_crit
print(t_reduced)
```

From the figure: $Z=0.88$

$Z\equiv\frac{P\bar{V}}{RT}$ so $\bar{V}=\frac{ZRT}{P}$

```{r}
z=0.88
molar_volume=z*t*r_barcm3molk/p
print(molar_volume)
```

We get $78.52 cm^3$ as the molar volume of pure water at 800C and 1000 bars.

## Question 2
Calculate $\bar{V}_{H_2O}$ and $f_{H_2O}^{P,T}$ at $T=800C$ and $P=1000 bars$ from the modified Redlich-Kwong equation of state. Use the following formulations for the attraction coefficient ($a$) and the excluded volume ($b$) from Bowers and Helgeson (1983) where T is in C.

```{r}
t_c=800
a=10^6*(111.3+50.7*exp(-0.983*10^(-2)*t_c))
b=14.6

```


## Question 3
We are looking for equilibrium conditions in P,T space, so we are going to try to find the conditions where $\Delta\bar{G}=0$

$$0=\Delta\bar{G}^\circ(P,T)=\Delta(\bar{H}^\circ)^{1,298}-T_{eq}\Delta(\bar{S}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\Delta\bar{C_P}^\circ dT-T_{eq}\displaystyle\int^{T_{eq}}_{298}\frac{\Delta\bar{C_P}^\circ}{T}dT+\displaystyle\int^{P_{eq}}_1\Delta\bar{V}^\circ dP$$

But unlike before, we need to break the volume term down into the solid and fluid components as these behave differently:$0=\Delta\bar{G}^\circ(P,T)=\Delta(\bar{H}^\circ)^{1,298}-T_{eq}\Delta(\bar{S}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\Delta\bar{C_P}^\circ dT-T_{eq}\displaystyle\int^{T_{eq}}_{298}\frac{\Delta\bar{C_P}^\circ}{T}dT+\displaystyle\int^{P_{eq}}_1\Delta\bar{V_s}^\circ dP+\displaystyle\int^{P_{eq}}_1\nu_f\bar{V_f}dP$

From the last problem set we know that the slope doesn't change with $P$ or $T$ so:

$\Delta(\bar{H}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\Delta\bar{C_P}^\circ dT\approx\Delta(\bar{H}^\circ)^{1,T}$

$\Delta(\bar{S}^\circ)^{1,298}+\displaystyle\int^{T_{eq}}_{298}\frac{\Delta\bar{C_P}^\circ}{T}dT\approx \Delta(\bar{S}^\circ)^{1,T}$

$\frac{dP}{dT}=\frac{\Delta\bar{S}}{\Delta\bar{V}}$

Once we substitute these in we have:

$\Delta(\bar{G}^\circ)^{P,T}=\Delta(\bar{H}^\circ)^{1,T}-T\Delta(\bar{S}^\circ)^{1,T}+\Delta\bar{V_s}^\circ(P-P_r)+\displaystyle\int^{P_{eq}}_1\nu_f\bar{V_f}^\circ dP$

Now we need to deal with the volume term for our fluids. So our only complication is fugacity. For an ideal gas: $(f_i^\circ)^{P,T}=P$ so $\displaystyle\int^P_1\bar{V_f}^\circ dP=\nu_fRTln(f_f^\circ)^{P,T}$

Using a table of fugacities how can we do this?

$$\Delta(\bar{H}^\circ)^{1,T}-T\Delta(\bar{S}^\circ)^{1,T}+\Delta\bar{V_s}^\circ(P-P_r)+\nu_fRTln(f_f^\circ)^{P,T}=0$$

$\nu_f=1$ so we can ignore that.

Now the plan of attack (shown explicitly once in the code):

1. pick a $T$ to start
2. pick a $P$ which sets $(f_i^\circ)^{P,T}$ from the table
3. evaluate $\Delta(\bar{G}^\circ)^{P,T}$
4. Is $\Delta(\bar{G}^\circ)^{P,T}>0 \rightarrow$ choose a lower pressure, if $\Delta(\bar{G}^\circ)^{P,T}<0\rightarrow$ choose a higher pressure
5. iterate

```{r}
h=24184 #cal
s=40.817 #cal K-1
delta_barv_solid=-0.113 #cal bar-1
p_ref<-1
```

First we have to figure out what temperature to start at. We need to find the lowest temperature where the lowest pressure and fugacity give a negative $(\Delta\bar{G}^\circ)^{P,T}$

```{r}
t<-200+273.15
p<-100
f<-15
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this one is really high, but it's already at the lowest pressure we have data for so we are going to skip this temperature
t<-300+273.15
f<-68
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #still not an option
t<-400+273.15
f<-87
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #still not an option
t<-500+273.15
f<-92
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # woo we have a negative! Let's check the temperature above this to see if it was negative too:
t<-480+273.15
f<-92
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) # nope, so we'll start at T=500
```

Alright so the lowest temperature we care about is 500C, so let's start figuring out the pressure conditions that provide us with equilibrium:

```{r}
t<-500+273.15
p<-100
f<-92
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #we checked this before, now let's check the next pressure
p<-200
f<-170
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 100 and 200 bars
#onto the next temperature
t<-520+273.15
p<-200
f<-173
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-300
f<-240
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 200 and 300 bars
t<-540+273.15
p<-300
f<-245
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-400
f<-305
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 300 and 400 bars
t<-560+273.15
p<-500
f<-369
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-600
f<-417
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 500 and 600 bars
t<-580+273.15
p<-900
f<-562
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-1000
f<-601
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 900 and 1000 bars
t<-600+273.15
p<-1400
f<-794
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-1500
f<-834
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 1400 and 1500 bars
t<-620+273.15
p<-2000
f<-1109
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-2100
f<-1152
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 2000 and 2100 bars
t<-640+273.15
p<-2800
f<-1603
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-2900
f<-1663
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 2800 and 2900 bars
t<-660+273.15
p<-3600
f<-2236
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-3700
f<-2312
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 3600 and 3700 bars
t<-680+273.15
p<-4400
f<-3035
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-4500
f<-3130
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 4400 and 4500 bars
t<-700+273.15
p<-5300
f<-4162
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-5400
f<-4285
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 5300 and 5400 bars
t<-720+273.15
p<-6200
f<-5583
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-6300
f<-5739
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 2800 and 2900 bars
t<-740+273.15
p<-7200
f<-7560
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-7300
f<-7760
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 2800 and 2900 bars
t<-760+273.15
p<-8100
f<-9834
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is negative, so onto the next highest pressure
p<-8200
f<-10082
print(h-t*s+delta_barv_solid*(p-p_ref)+r_calmolk*t*log(f)) #this is positive, so the actual pressure of equilibrium is between 8100 and 8200 bars
t<-c(500,520,540,560,580,600,620,640,660,680,700,720,740,760)
t<-t+273.15
lower_p<-c(100,200,300,500,900,1400,2000,2800,3600,4400,5300,6200,7200,8100)
upper_p<-c(200,300,400,600,1000,1500,2100,2900,3700,4500,5400,6300,7300,8200)

```
## Question 4


```{r}
h=23019 #cal
s=38.334 #cal k-1
delta_barv_solid=-0.471 #cal bar-1
t<-c(350,400,450,500,550,600,650,700)
t<-t+273.15

p<-1000
f<-1096
t_term<-t[4]*s
delv_term<-delta_barv_solid*(p-p_ref)
fug_term<-r_calmolk*t[1]*log(f)
test1<-h-t[6]*s+delta_barv_solid*(p-p_ref)+r_calmolk*t[6]*log(f)
```
## Question 5
